---
title: "Time Series (732A62) Lab1"
author: "Anubhav Dikshit(anudi287) and Maximilian Pfundstein(maxpf364)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    pdf_document:
      toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)

library("ggplot2")
library("tidyverse")
library("gridExtra") # combine plots
library("knitr")


# The palette with black:
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
                "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
set.seed(12345)
```

\newpage

# Assignment 1. Computations with simulated data


\fbox{\begin{minipage}{46.7em}
a) Generate two time series $x_t = -0.8x_{t-2} + w_t,$ where $x_0 = x_1 = 0$ and $x_t= cos(\frac{2 \pi t}{5}) $ with 100 observations each. Apply a smoothing filter $v_t = 0.2(x_t + x_{t-1} + x_{t-2} + x_{t-3} + x_{t-4})$ to these two series and compare how the filter has affected them.
\end{minipage}}

```{r}
set.seed(12345)

n = 100
x <- vector(length = n)
x2 <- vector(length = n)

x[1] <- 0
x[2] <- 0

#first series generation
for(i in 3:n){
  x[i] <- -0.8 * x[i-2] + rnorm(1,0,1)
}

#second series generation
for(i in 1:n){
  x2[i] <- cos(0.4*pi*i)
}

# smoothing filter function
smoothing_filter <- function(x){
v <- vector(length = length(x))
for(i in 5:length(x)){
  v[i] = 0.2 * (x[i] + x[i-1] + x[i-2] + x[i-3] + x[i-4])
}
return(v)
}

#generate smoothed series
smooth_x <- smoothing_filter(x)
smooth_x2 <- smoothing_filter(x2)

#adding everything to a dataframe
df <- cbind(x,x2,smooth_x,smooth_x2) %>% as.data.frame() %>% mutate(index=1:100)

ggplot(df, aes(x=index)) + 
  geom_line(aes(y=x, color="Original Time Series")) + 
  geom_line(aes(y=smooth_x, color="Smoothened Time Series")) + 
  ggtitle("Plot of 1st time series and its smoothened version") +
    scale_colour_manual("", breaks = c("Original Time Series", "Smoothened Time Series"),
                        values = c("#CC79A7", "#000000"))

ggplot(df, aes(x=index)) + 
  geom_line(aes(y=x2, color="Original Time Series")) + 
  geom_line(aes(y=smooth_x2, color="Smoothened Time Series")) + 
  ggtitle("Plot of 2nd time series and its smoothened version") +
    scale_colour_manual("", breaks = c("Original Time Series", "Smoothened Time Series"),
                        values = c("#CC79A7", "#000000"))

```

\fbox{\begin{minipage}{46.7em}
b) Consider time series $x_t - 4 x_{t-1} + 2 x_{t-2} + x_{t-5} = w_t + 3 w_{t-2} + w_{t-4} - 4w_{t-6}$. Write an appropriate R code to investigate whether this time series is casual and invertible.
\end{minipage}}


A linear process $X_t$ is causal(strictly, a causal function of $w_t$) if there is a 
$$\psi(B) = \psi_0 + \psi_1 B + \psi_2 B^2 +....  $$
with $$ \sum^\infty_{j=0} |\psi_j| < \infty $$
and
$$ X_t = \psi(B) w_t$$
```{r}
set.seed(12345)
w <- rnorm(100,0,1)
x <- rep(0,100)

for(i in 7:100){
x[i] = w[i] + 3 * w[i-2] + w[i-4] - 4 * w[i-6] + 4 * x[i-1] - 2 * x[i-2] - x[i-5]   
}
plot(x)

```

\fbox{\begin{minipage}{46.7em}
c) Use built-in R functions to simulate 100 observations from the process $ x_t + \frac{3}{4} x_{t-1} = w_t - \frac{1}{9} w_{t-2},$ compute sample ACF and theoretical ACF, use seed 54321. Compare the ACF plots.
\end{minipage}}

```{r}
set.seed(54321)

```

# Appendix

```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```


